---
title: "Confidence Interval for Population Proportion"
output:
  tufte::tufte_handout: default
  tufte::tufte_html: default
---

```{r, echo=FALSE, warning=FALSE}
load('data/brfss2019mo.RData')
```

# What it is

In general, confidence intervals use observed data to give a range of values where the population parameter is thought to exist. Confidence intervals for $p$ use the sample proportion $\hat{p}$ and the sample size $n$ to build an interval for $p$ from the binomial distribution.


# When to use it

The confidence interval for $p$ requires the conditions of a binomial distribution, which include a sample of a fixed and known size, each observation an independent binary outcome. Many textbooks teach the normal model approximation method, which requires a large sample size (usually defined as at least 10 success and at least 10 failures); however, the `binom.test()` R function used here does not employ the normal approximation and will work for any sample size.


# How to use it

In R, the confidence interval for $p$ is obtained most easily through the `binom.test()` function. Suppose we want to estimate the prevalence of diabetes among Missouri adults. First, we need to count the number of diabetic respondents and the total sample size:

```{r}
table(mydat$diabetes, useNA='ifany')
```

So, there are 264 positive responses out of 1998 (= 264 + 1734) valid responses. Then, we can use these values as inputs to `binom.test()`. Appending `$conf.int` will return only the confidence interval^[See *Hypothesis Test for Population Proportion* for more detail about the output of `binom.test()`]:

```{r}
binom.test(x=264, n=1998, conf.level=0.95)$conf.int
```

So, from our data we are 95% confident that somewhere between 12% and 15% of the Missouri adults have diabetes.

We can change the confidence level by changing `conf.level` in the R code. For example, using `conf.level=0.90` will give a 90% confidence interval.


# Why it works

R uses a highly computational process to obtain the the 95% confidence interval above. For large samples (at least 10 success and at least 10 failures), the Central Limit Theorem  suggests the $Z$ model can be used instead. In particular,

$$
\frac{\hat{p} - p}{\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}} \sim Z
$$

Since there is about a 0.95 probability that an observation will be within two standard deviations of the distribution mean^[And the $Z$ distirbution has a standard deviation of 1], we have:

$$
P\left( -2 < \frac{\hat{p} - p}{\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}} < 2 \right) = 0.95
$$

Then we can solve the inequality in the probability to obtain just $p$ on the inside^[Steps: (1) multiply by $\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}$; (2) subtract $\hat{p}$; (3) multiply by $-1$, remembering to switch the direction of the inequality signs]:

$$
P\left(\hat{p} - 2\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}} < p < \hat{p} + 2\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}\right) = 0.95
$$

So, there is a 95% chance that $p$ is between $\hat{p} - 2\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}$ and $\hat{p} + 2\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}$. Making the result a bit more compact, the normal approximation 95% confidence interval for $p$ is

$$
\hat{p} \pm 2 \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}
$$
