---
title: "Normal model"
output:
  tufte::tufte_handout: default
  tufte::tufte_html: default
date: 'Date: `r format(Sys.time(), "%Y-%m-%d")`'
---

# What it is

```{r, echo=FALSE, fig.margin=TRUE, fig.cap="A normal model with mean 0 and standard deviation 1."}
par(mar=c(2, 1, 1, 1))
curve(dnorm, from=-4, to=4, lwd=2, 
      xlab='', 
      yaxt='n', ylab='', frame=FALSE)
abline(h=0, col='grey')
```

The normal model is the ubiquitous "bell-shaped curve" that is one example of a model for continuous data. The normal model is symmetric and peaks at the mean.


# When to use it

The normal model is a common one in statistical inference. The CLT shows that for large samples, the totals and averages of observations follow approximately a normal model (see *Why it works*). In addition, many biological and psychological measurements (height, IQ, test scores) follow normal models.


# How to use it

The **Empirical Rule** can be useful to make quick estimates. It states that for a normal model, 68% of the probability falls within 1 standard deviation of the mean, 95% falls within 2 standard deviations of the mean, and 99.7% falls within three standard deviations of the mean.

To compute normal probabilities in R, use the `pnorm()` function. If $Y$ is modeled by a normal model with mean $m$ and standard deviation $s$, then $P(Y\leq y)=$ `pnorm(q=y, mean=m, sd=s)`. Multiple `pnorm`s can be used to find other types of areas. For example, say the mean is $m=100$ and standard deviation is $s=15$^[This is approximately the distribution of IQ scores in adults]. Then 

```{r, echo=FALSE, fig.margin=TRUE, fig.cap="P(90 < Y < 120) = P(Y < 120) - P(Y < 90)."}
par(mar=c(2, 1, 1, 1))
curve(dnorm(x, mean=100, sd=15), from=60, to=140, lwd=2, 
      xlab='', 
      yaxt='n', ylab='', frame=FALSE)
x1 = seq(60, 120, 0.01)
y1 = dnorm(x1, mean=100, sd=15)
polygon(x=c(60,x1,120), y=c(0,y1,0), col='lightblue', border='slateblue', lwd=2)
x2 = seq(60, 90, 0.01)
y2 = dnorm(x2, mean=100, sd=15)
polygon(x=c(60,x2,90), y=c(0,y2,0), col='white', border='transparent', density=10, angle=-45, lwd=2)
abline(h=0, col='grey')
```

```{r}
# P(90 < Y < 120) 
pnorm(q=120, mean=100, sd=15) - pnorm(q=90, mean=100, sd=15)

# P(Y > 130)
1 - pnorm(q=130, mean=100, sd=15)
```


# Why it works

